---
title: "Chap_4"
author: "Dr K"
date: "June 27, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(MASS)
```

## Too much is too much
The first example presented in Chapter 1 applied the plot() function to a data frame, yielding an array of scatterplots with one for each pair of columns in the data frame. Thus, the number of plots in this array is equal to the square of the number of columns in the data frame.

This means that if we apply the plot() function to a data frame with many columns, we will generate an enormous array of scatterplots, each of which will be too small to be useful. The purpose of this exercise is to provide a memorable example.

Instructions
100 XP
Applying the plot() function to a data frame with k columns will create a k by k array of scatterplots.

Use the ncol() function to compute the number of plots in this array for the Cars93 data frame. Just print the result to the console.
Call plot() on Cars93 to generate the scatterplot array. Can you verify that your calculation was correct?

```{r}
# Compute the number of plots to be displayed
ncol(Cars93)^2

# Plot the array of scatterplots
plot(Cars93)
```

## Deciding how many scatterplots is too many
As noted in the video, the matplot() function can be used to easily generate a plot with several scatterplots on the same set of axes. By default, the points in these scatterplots are represented by the numbers 1 through n, where n is the total number of scatterplots included, but most of the options available with the plot() function are also possible by specifying the appropriate arguments.

This exercise asks you to set up a plot array with four of these multiple scatterplot displays, each including one more scatterplot than the previous one. The point of this exercise is to encourage you to judge for yourself how many scatterplots is too many?.

Instructions
100 XP
Instructions
100 XP
Run the first line of sample code provided to construct a character vector called keep_vars that names six variables from the UScereal data frame.
Using this vector of variable names, extract the data frame df containing only these variables from the UScereal data frame.
Use the par() function to set up the mfrow parameter to generate a two-by-two plot array.
Use the matplot() function to construct a two-scatterplot display of
protein and fat versus calories. Give this plot the title "Two scatterplots" using the title() function. Label the x-axis "calories" and the y-axis "" to get rid of the default y-axis label.
protein, fat and fibre versus calories. Give this plot the title "Three scatterplots". Similarly, label the x-axis "calories" and the y-axis "".
protein, fat, fibre and carbo versus calories. Give this plot the title "Four scatterplots". Again, label the x-axis "calories" and the y-axis "".
protein, fat, fibre, carbo, and sugars versus calories. Give this plot the title "Five scatterplots". Be sure to label the x-axis "calories" and the y-axis "" again.

```{r}
# Construct the vector keep_vars
keep_vars <- c("calories", "protein", "fat",
               "fibre", "carbo", "sugars")

# Use keep_vars to extract the desired subset of UScereal
df <- UScereal[, keep_vars]

# Set up a two-by-two plot array
par(mfrow = c(2, 2))

# Use matplot() to generate an array of two scatterplots
matplot(df$calories, df[, c("protein", "fat")], 
        xlab = "calories", ylab = "")

# Add a title
title("Two scatterplots")

# Use matplot() to generate an array of three scatterplots
matplot(df$calories, df[, c("protein", "fat", "fibre")], 
        xlab = "calories", ylab = "")

# Add a title
title("Three scatterplots")

# Use matplot() to generate an array of four scatterplots
matplot(df$calories, 
        df[, c("protein", "fat", "fibre", "carbo")], 
        xlab = "calories", ylab = "")

# Add a title
title("Four scatterplots")

# Use matplot() to generate an array of five scatterplots
matplot(df$calories, 
        df[, c("protein", "fat", "fibre", "carbo", "sugars")], 
        xlab = "calories", ylab = "")

# Add a title
title("Five scatterplots")
```

## How many words is too many?
The main point of the previous two exercises has been to show that scatterplot arrays lose their utility if they are allowed to become too complex, either including too many plots, or attempting to include too many scatterplots on one set of axes. More generally, any data visualization loses its utility if it becomes too complex.

This exercise asks you to consider this problem with wordclouds, displays that present words in varying sizes depending on their frequency. That is, more frequent words appear larger in the display, while rarer words appear in a smaller font.

In R, wordclouds are easy to generate with the wordcloud() function in the wordcloud package. This function is called with a character vector of words, and a second numerical vector giving the number of times each word appears in the collection used to generate the wordcloud.

Two other useful arguments for this function are scale and min.freq. The scale argument is a two-component numerical vector giving the relative size of the largest word in the display and that of the smallest word. The wordcloud only includes those words that occur at least min.freq times in the collection and the default value for this argument is 3.

Instructions
100 XP
Instructions
100 XP
The wordcloud package is already loaded in your workspace.

Use the table() function to create the variable mfr_table that tabulates the number of times each level of the Manufacturer variable appears in the Cars93 data frame.
Use the wordcloud() function to display this manufacturer data, setting the scale argument to c(2, 0.25) to make the wordcloud fit in the display window.
By default, the wordcloud() function shows only those words that appear 3 or more times. Use the min.freq argument to obtain a display with all Manufacturer values in the wordcloud.
Use the table() function to create the variable model_table that tabulates the number of times each level of the Model variable appears in the Cars93 data frame.
Use the wordcloud() function with the scale argument set to c(0.75, 0.25) and the min.freq argument set to display all Model values. Use the zoom feature in the 'Plots' window to see this wordcloud more clearly. Does it convey useful information?  

```{r}
require(wordcloud) #dont include in datacamp answer
# Create mfr_table of manufacturer frequencies
mfr_table <- table(Cars93$Manufacturer)

# Create the default wordcloud from this table
wordcloud(words = names(mfr_table), 
          freq = as.numeric(mfr_table), 
          scale = c(2, 0.25))

# Change the minimum word frequency
wordcloud(words = names(mfr_table), 
          freq = as.numeric(mfr_table), 
          scale = c(2, 0.25), 
          min.freq = 1)

# Create model_table of model frequencies
model_table <- table(Cars93$Model)

# Create the wordcloud of all model names with smaller scaling
wordcloud(words = names(model_table), 
          freq = as.numeric(model_table), 
          scale = c(0.75, 0.25), 
          min.freq = 1)
```

## The Anscombe quartet
This exercise and the next one are based on the Anscombe quartet, a collection of four datasets that appear to be essentially identical on the basis of simple summary statistics like means and standard deviations. For example, the mean x-values for these datasets are identical to three digits, while the mean y-values differ only in the third digit.

In spite of these apparent similarities, the behavior of the four datasets is quite different and this becomes immediately apparent when we plot them.

Instructions
100 XP
Use the par() function to set up a two-by-two plot array.
Using plot(), create 4 separate plots from the anscombe data frame:
y1 vs. x1
y2 vs. x2
y3 vs. x3
y4 vs. x4
  
```{r}
# Set up a two-by-two plot array
par(mfrow = c(2, 2))

# Plot y1 vs. x1 
plot(anscombe$x1, anscombe$y1)

# Plot y2 vs. x2
plot(anscombe$x2, anscombe$y2)

# Plot y3 vs. x3
plot(anscombe$x3, anscombe$y3)

# Plot y4 vs. x4
plot(anscombe$x4, anscombe$y4)
```
  
## The utility of common scaling and individual titles
The plots you generated in the previous exercise showed that the four Anscombe quartet datasets have very different appearances, but a careful examination of these plots reveals that they exhibit different ranges of x and y values.

The point of this exercise is to illustrate how much more clearly we can see the differences in these datasets if we plot all of them with the same x and y ranges. This exercise also illustrates the utility of improving the x- and y-axis labels and of adding informative plot titles.

Instructions
100 XP
Examine the range of x and y values from the previous four plots to determine a common range that covers the ranges of all four of these plots.
Set up a two-by-two plot array using the par() function.
Plot the y1 variable against the x1 variable using these common ranges, with x-axis label "x value" and y-axis label "y value", and use the main argument to add the title "First dataset".
Repeat for the other three Anscombe data pairs, adding the appropriate titles.
  
```{r}
# Define common x and y limits for the four plots
xmin <- min(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4)
xmax <- max(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4)
ymin <- min(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
ymax <- max(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)

# Set up a two-by-two plot array
par(mfrow = c(2, 2))

# Plot y1 vs. x1 with common x and y limits, labels & title
plot(anscombe$x1, anscombe$y1,
     xlim = c(xmin, xmax),
     ylim = c(ymin, ymax),
     xlab = "x value", ylab = "y value",
     main = "First dataset")

# Do the same for the y2 vs. x2 plot
plot(anscombe$x2, anscombe$y2,
     xlim = c(xmin, xmax),
     ylim = c(ymin, ymax),
     xlab = "x value", ylab = "y value",
     main = "Second dataset")

# Do the same for the y3 vs. x3 plot
plot(anscombe$x3, anscombe$y3,
     xlim = c(xmin, xmax),
     ylim = c(ymin, ymax),
     xlab = "x value", ylab = "y value",
     main = "Third dataset")

# Do the same for the y4 vs. x4 plot
plot(anscombe$x4, anscombe$y4,
     xlim = c(xmin, xmax),
     ylim = c(ymin, ymax),
     xlab = "x value", ylab = "y value",
     main = "Fourth dataset")
```

## Using multiple plots to give multiple views of a dataset
As noted in the video, another useful application of multiple plot arrays besides comparison is presenting multiple related views of the same dataset.

This exercise illustrates this idea, giving four views of the same dataset: a plot of the raw data values themselves, a histogram of these data values, a density plot, and a normal QQ-plot.

Instructions
100 XP
Note that the MASS and car packages have been pre-loaded, making the geyser data and the truehist() and qqPlot() functions available for your use.
Use the par() function to set the mfrow parameter for a two-by-two plot array.
In the upper left, use the plot() function to show the values of the duration variable from the geyser dataset, using the main argument to specify the plot title as "Raw data".
In the upper right, use the truehist() function to generate a histogram of the duration data, using the main argument to give this plot the title "Histogram".
In the lower left, use the plot() and density() functions to display the density of the duration values, using the main argument to give this plot the title "Density".
In the lower right, use the qqPlot() function from the car package to display a normal QQ-plot of the duration data, using the main argument to give this plot the title "QQ-plot".  

```{r}
require(car) #dont include in datacamp answer
# Set up a two-by-two plot array
par(mfrow = c(2, 2))

# Plot the raw duration data
plot(geyser$duration, main = "Raw data")

# Plot the normalized histogram of the duration data
truehist(geyser$duration, main = "Histogram")

# Plot the density of the duration data
plot(density(geyser$duration), main = "Density")

# Construct the normal QQ-plot of the duration data
qqPlot(geyser$duration, main = "QQ-plot")
```

## Constructing and displaying layout matrices
The video illustrated how to set up a layout matrix to be used with the layout() function in creating a plot array. You can think of the layout matrix as the plot pane, where a 0 represents empty space and other numbers represent the plot number, which is determined by the sequence of visualization function calls. For example, a 1 in the layout matrix refers to the visualization that was first called, a 2 refers to the plot of the second visualization call, etc. This exercise asks you to create your own 3 x 2 layout matrix, using the c() function to concatenate numbers into vectors that will form the rows of the matrix.
  


You will then use the matrix() function to convert these rows into a matrix and apply the layout() function to set up the desired plot array. The convenience function layout.show() can then be used to verify that the plot array has the shape you want.

As reference, feel free to use the slides from the video, which you can access by clicking on the "Slides" tab next to the "R Console" tab. Pay close attention to the example layout matrix and the resulting plot array.

Instructions
100 XP
Using the matrix() function, create a matrix called layoutMatrix with three rows and two columns:
the first row designates an empty plot to the left and plot 1 to the right.
the second row designates plot 2 to the left and an empty plot to the right.
the third row designates an empty plot to the left and plot 3 to the right.
Use the layout() function to set up the desired plot array.
Use the layout.show() function to show the arrangement of all three plots.  


```{r}
# Use the matrix function to create a matrix with three rows and two columns
layoutMatrix <- matrix(
  c(
    0, 1,
    2, 0,
    0, 3
  ), 
  byrow = TRUE, 
  nrow = 3
)
# Call the layout() function to set up the plot array
layout(layoutMatrix)

# Show where the three plots will go 
layout.show(3)
```  

## Creating a triangular array of plots
The previous exercise asked you to create a plot array using the layout() function. Recall the layout matrix from the previous exercise:

> layoutMatrix
     [,1] [,2]
[1,]    0    1
[2,]    2    0
[3,]    0    3
This exercise asks you to use this array to give three different views of the whiteside data frame.

The first plot, on the upper right of the plot array, shows the relationship of Gas and Temp using all data from whiteside. The second plot, in the center left of the plot array, shows the relationship of the two variables using data where Insul is equal to "Before". Finally, the third plot, on the lower right of the plot array, shows the relationship using data where Insul is equal to "After".

The primary motivation for this exercise is that it is not possible to construct a plot array in this format using the mfrow parameter, since the array is not rectangular.

Instructions
100 XP
Instructions
100 XP
The layout matrix you set up in the previous is available in your workspace as layoutMatrix. The whiteside data frame is already available in your workspace as well.

Call the layout() function on layoutMatrix to set up the plot array.
Construct a vector indexB that points to only those records with the Insul value "Before" and a vector indexA that points to only those records with the Insul value "After".
Plot the Gas versus Temp values from the indexB data in the upper right plot in your array, using the y-axis limits c(0, 8). Give this plot the title "Before data only".
Plot the Gas versus Temp values from the complete dataset in the center left plot, using the same y-axis limits as the first plot. Give this plot the title "Complete dataset".
Plot the Gas versus Temp values from the indexA data in the lower right plot, again using the same y-axis limits. Give this plot the title "After data only".
  
```{r}
# Set up the plot array
layout(layoutMatrix)

# Construct the vectors indexB and indexA
indexB <- which(whiteside$Insul == "Before")
indexA <- which(whiteside$Insul == "After")

# Create plot 1 and add title
plot(whiteside$Temp[indexB], whiteside$Gas[indexB],
     ylim = c(0, 8))
title("Before data only")

# Create plot 2 and add title
plot(whiteside$Temp, whiteside$Gas,
     ylim = c(0, 8))
title("Complete dataset")

# Create plot 3 and add title
plot(whiteside$Temp[indexA], whiteside$Gas[indexA],
     ylim = c(0, 8))
title("After data only")
```
  
## Creating arrays with different sized plots
Besides creating non-rectangular arrays, the layout() function can be used to create plot arrays with different sized component plots -- something else that is not possible by setting the par() function's mfrow parameter.

This exercise illustrates the point, asking you to create a standard scatterplot of the zn versus rad variables from the Boston data frame as a smaller plot in the upper left, with a larger sunflower plot of the same data in the lower right.

Instructions
100 XP
Instructions
100 XP
The Boston data frame in the MASS package is already available in your workspace.

Using the c() function:
Create the three-element vector row1 with values (1, 0, 0).
Create the three-element vector row2 with values (0, 2, 2).
Combine the first vector with two copies of the second vector into layoutVector, a vector of length 9.
Using the matrix() function, convert layoutVector into the 3-by-3 matrix layoutMatrix whose first row is row1 and whose second and third rows are row2.
Use the layout() function with layoutMatrix to configure a two-element plot array.
In the first (smaller) plot, create a standard scatterplot of the zn variable versus the rad variable from the Boston data frame.
In the second (larger) plot, create a sunflower plot using the sunflowerplot() function and the same variables.
  
```{r}
# Create row1, row2, and layoutVector
row1 <- c(1, 0, 0)
row2 <- c(0, 2, 2)
layoutVector <- c(row1, rep(row2, 2))

# Convert layoutVector into layoutMatrix
layoutMatrix <- matrix(layoutVector, byrow = TRUE, nrow = 3)

# Set up the plot array
layout(layoutMatrix)

# Plot scatterplot
plot(Boston$rad, Boston$zn)

# Plot sunflower plot
sunflowerplot(Boston$rad, Boston$zn)
```

