---
title: "Chap_5.rmd"
author: "Dr K"
date: "July 2, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Some plot functions also return useful information
As shown in the video, calling the barplot() function has the side effect of creating the plot we want, but it also returns a numerical vector with the center positions of each bar in the plot. This value is returned invisibly so we don't normally see it, but we can capture it with an assignment statement.

These return values can be especially useful when we want to overlay text on the bars of a horizontal barplot. Then, we capture the return values and use them as the y parameter in a subsequent call to the text() function, allowing us to place the text at whatever x position we want but overlaid in the middle of each horizontal bar. This exercise asks you to construct a horizontal barplot that exploits these possibilities.

Feel free to reference a similar example in the slides by clicking on the "Slides" tab next to the "R Console" tab.



Instructions
100 XP
Instructions
100 XP
The Cars93 data frame from the MASS package is already available in your workspace.

Use the table() function to generate the object tbl summarizing the number of records listing each value of the Cylinders variable from the Cars93 data frame.
Use the barplot() function with the horiz argument set to TRUE to construct a horizontal barplot of this record summary, specifying the color of the bars to be "transparent" and returning the vector mids giving the vertical positions of the centers of each bar in the plot. Specify the names.arg argument as "" to suppress the y-axis legend on this plot.
Use the text() function to label the Cylinders value for each bar in the barplot at the horizontal position 20. The names() function may prove useful here.
Use the text() function to list the counts for each cylinders value at the horizontal position 35. Here, the as.numeric() function may prove useful.

```{r}
require(MASS) #Dont include in datacamp answer
# Create a table of Cylinders frequencies
tbl <- table(Cars93$Cylinders)

# Generate a horizontal barplot of these frequencies
mids <- barplot(tbl, horiz = TRUE, 
                col = "transparent",
                names.arg = "")

# Add names labels with text()
text(20, mids, names(tbl))

# Add count labels with text()
text(35, mids, as.numeric(tbl))
```  

## Using the symbols() function to display relations between more than two variables
The scatterplot allows us to see how one numerical variable changes with the values of a second numerical variable. The symbols() function allows us to extend scatterplots to show the influence of other variables.

This function is called with the variables x and y that define a scatterplot, along with another argument that specifies one of several possible shapes. Here, you are asked to use the circles argument to create a bubbleplot where each data point is represented by a circle whose radius depends on the third variable specified by the value of this argument.

Instructions
100 XP
Instructions
100 XP
The Cars93 data frame from the MASS package is already available in your workspace.

#### Hint  

Use the symbols() function with its default settings and the appropriate arguments to create a bubble plot of MPG.city versus Horsepower from the Cars93 data frame, with the bubble area by the Price variable. Note that this means the bubble radius should be proportional to the square root (sqrt()) of Price.
Re-create the first plot, but with the optional argument inches set to 0.1.

Like plot(), the x and y arguments for the symbols() function (which are the first two arguments if no argument names are given) determine the x- and y-position of each data point. The other arguments for symbols() determine what shape these points are, sizing them with other data variables.
In this exercise, the optional argument circles is used to request a bubbleplot and specify the size of the bubbles by giving a variable name.
The optional argument inches specifies the maximum symbol size; here, the default value is inches = 1 and this is too large.  

```{r}
require(Cars93) #Don't include in datacamp answer
# Call symbols() to create the default bubbleplot
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price))

# Repeat, with the inches argument specified
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price),
        inches = 0.1)
```
  
## Saving plot results as files
In an interactive R session, we typically generate a collection of different plots, often using the results to help us decide how to proceed with our analysis. This is particularly true in the early phases of an exploratory data analysis, but once we have generated a plot we want to share with others, it is important to save it in an external file.

R provides support for saving graphical results in several different external file formats, including jpeg, png, tiff, or pdf files. In addition, we can incorporate graphical results into external documents, using tools like the Sweave() function or the knitr package. One particularly convenient way of doing this is to create an R Markdown document, an approach that forms the basis for another DataCamp course.

Because png files can be easily shared and viewed as e-mail attachments and incorporated into many slide preparation packages (e.g. Microsoft Powerpoint), this exercise asks you to create a plot and save it as a png file. The basis for this process is the png() function, which specifies the name of the png file to be generated and sets up a special environment that captures all graphical output until we exit this environment with the dev.off() command.

Instructions
100 XP
Use the png() function to direct all subsequent plot results to the external file bubbleplot.png.
Run the code to re-create the second bubble plot from the previous exercise.
Exit the png() environment to return graphics control to your session by calling dev.off().
Use the list.files() code provided to verify that bubbleplot.png was created. To do this, you need to specify "png" as the value to the pattern argument to list.files().

#### Hint

Call the png() function with the filename "bubbleplot.png" to save subsequent plot results in this file.
Next, call the symbols() function as you did in the previous exercise to create a bubbleplot.
Finally, call dev.off() with no arguments to stop saving all graphics output to the external file bubbleplot.png.
The call list.files(pattern = "xxx") returns a list of all external file names containng the pattern "xxx". 

```{r}
# Call png() with the name of the file we want to create
png("bubbleplot.png")

# Re-create the plot from the last exercise
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price),
        inches = 0.1)

# Save our file and return to our interactive session
dev.off()

# Verify that we have created the file
list.files(pattern = "png")
```

## Iliinsky and Steele's 12 recommended colors
This exercise asks you to create a horizontal barplot that shows Iliinsky and Steele's set of recommended 12 colors, in descending order of desirability from the top of the plot to the bottom. Also, the first six "more preferred" colors are displayed with longer bars to visually emphasize their preferred status over the other six.

Code is provided to create the character vector IScolors with the names of the 12 colors recommended by Iliinsky and Steele, in their recommended order. We've also created the numeric vector barWidths containing the value 2 for the first six Iliinsky and Steele colors and the value 1 for the next six.

In this exercise, you'll use the barplot() function to create the horizontal barplot shown in the preceeding figure, with the color names to the left of the bars and each bar drawn in the indicated color.

Instructions
100 XP
Notice how the barWidths vector of length 12 contains the length of each bar shown.
The 6 longer bars have a value of 2.
The 6 shorter ones have a value of 1.
Recreate the horizontal barplot you see on the right, using the horiz, col, axes, names.arg, and las arguments to barplot(). Note that since the horizontal barplot is constructed from the bottom up, it is necessary to reverse the IScolors and barWidths vectors in the barplot() function call, using the rev() function.

#### Hint

Hint
Pass rev(barWidths) to the barplot() function and set the names.arg and col arguments equal to rev(IScolors). Also recall that the axes and horiz arguments take boolean values.

```{r}
# Iliinsky and Steele color name vector
IScolors <- c("red", "green", "yellow", "blue",
              "black", "white", "pink", "cyan",
              "gray", "orange", "brown", "purple")

# Create the data for the barplot
barWidths <- c(rep(2, 6), rep(1, 6))

# Recreate the horizontal barplot with colored bars
barplot(rev(barWidths), horiz = TRUE, 
        col = rev(IScolors), axes = FALSE,
        names.arg = rev(IScolors), las = 1)
```

## Using color to enhance a bubbleplot
In a previous exercise, you saw how the symbols() function could be used to generate a bubbleplot showing the relationship between three variables (specifically, Horsepower, MPG.city, and Price from the Cars93 data frame). There, the basic format was a scatterplot of MPG.city versus Horsepower, with points represented as bubbles, sized by Price.

Here, you are asked to create a variation on this plot, using the factor variable Cylinders to determine both the size and the color of the bubbles. To do this, note that as.numeric(Cars93$Cylinders) generates a sequence of numerical values from 1 to 6 that specify the six unique levels of the Cylinders factor.

The point of this exercise is to show that, in cases where color is an option, the clarity of this bubbleplot can be improved substantially through the use of color in addition to symbol size. Since the Cylinders variable exhibits six unique values, six colors are required to make this enhancement, so the top six colors recommended by Iliinsky and Steele are used here.

Instructions
100 XP
Instructions
100 XP
The Cars93 data frame in the MASS package is available in your workspace, and the vector IScolors of color names is created in the code included at the beginning of the exercise.

Create the vector cylinderLevels giving numerical labels to the unique values of the Cylinders variable.
Using the symbols() function, recreate the bubbleplot shown on the right of MPG.city vs. Horsepower with the bubbles sized by the cylinderLevels variable.
Specify a maximum circle radius of 0.2 inches.
Use the bg parameter to color each bubble according to the numerical level of the Cylinders factor variable (i.e., the cylinderLevels values), using the first six recommended colors from Iliinsky and Steele (e.g. the first Cylinder level should be colored "red", the second level colored "green", etc.)

#### Hint

Note that cylinderLevels = as.numeric(Cars93$Cylinders).
Recall that the circles argument determines the size of the bubbles.
The background color specification is bg = IScolors[cylinderLevels].

```{r}
# Iliinsky and Steele color name vector
IScolors <- c("red", "green", "yellow", "blue",
              "black", "white", "pink", "cyan",
              "gray", "orange", "brown", "purple")
# Create the `cylinderLevels` variable
cylinderLevels <- as.numeric(Cars93$Cylinders)


# Create the colored bubbleplot
symbols(Cars93$Horsepower, Cars93$MPG.city, 
        circles = cylinderLevels, inches = 0.2, 
        bg = IScolors[cylinderLevels])
```

## Using color to enhance stacked barplots
The most common barplot consists of a collection of vertical bars, each representing some characteristic of one of a finite number of datasets or data subsets. Several previous exercises have illustrated the utility of the somewhat less common horizontal barplot, useful in part because it allows text to be displayed horizontally across the bars, as illustrated in Exercise 1.

Another useful variant of the standard bar plot is the stacked bar plot, where each bar in the plot is partitioned into segments characterizing portions of the data characterized by the bar. Stacked bar plots can also be generated using the barplot() function. Here, each bar is specified by a matrix whose columns specify the heights of the segments in each bar.

By default, the barplot() function generates stacked bar plots using different shades of gray for the different segments of each bar in the plot. The point of this exercise is to show that, if we can use it, color can be a more effective alternative.

Instructions
100 XP
The character vector IScolors from the previous exercises is still available in your workspace.

Create the table tbl giving the record counts for each Cylinders value at each Origin value in the Cars93 dataset (this table should have 6 rows and 2 columns).
Using the barplot() function, create a stacked barplot that summarizes this information using shades of gray.
Recreate this first plot, but using the first six Iliinsky and Steele colors for the six Cylinders levels.

#### Hint

Pass the two variables into a single table() function to create tbl.
Note that calling barplot() with a two-dimensional table creates a stacked barplot.
The barplot() function has an argument called col for colors. Use the IScolors vector when you're specifying the colors in your second call to barplot().

```{r}
# Create a table of Cylinders by Origin
tbl <- table(Cars93$Cylinders, Cars93$Origin)

# Create the default stacked barplot
barplot(tbl)

# Enhance this plot with color
barplot(tbl, col = IScolors)
```

## The tabplot package and grid graphics
The tabplot package allows you to tap into the power of grid graphics without explicit knowledge of how the system works under the hood. The main function in tabplot is tableplot(), developed to visualize data distributions and relationships between variables in large datasets.

Specifically, the tableplot() function constructs a set of side-by-side horizontal barplots, one for each variable. This function works best when viewing up to about 10 variables at a time, for datasets with arbitrarily many records. In this exercise, you are asked to apply this function to a dataset with just under 68,000 records and 11 variables.

The tableplot() function is called with a data frame and, if no optional arguments are specified, it selects the first data column as the reference variable. This variable can be of any type, but the display is easiest to explain when it's numeric, as in the example considered here.

For further details, refer to the vignette Visualization of large datasets with tabplot that accompanies the help files for the tabplot package.

Instructions
100 XP
Load the insuranceData package.
Use the data() function to load the dataCar data frame.
Load the tabplot package the normal way, but surround your call to library() with the suppressPackageStartupMessages() function to avoid a bunch of unncessary output from printing to the console.
Apply the tableplot() function to the dataCar data frame.

#### Hint

Hint
Load the insuranceData package with the library() function.
To make the dataCar data frame available, use the data() function.
Load the tabplot package as suppressPackageStartupMessages(library(tabplot)).
Simply call tableplot() on dataCar after you've loaded the appropriate packages and data.  

```{r}
if(!require("insuranceData")) {install.packages("insuranceData");require("insuranceData")}
if(!require("tabplot")) {install.packages("tabplot");require("tabplot")}
#Dont include lines above in datacamp answer

# Load the insuranceData package
library(insuranceData)

# Use the data() function to load the dataCar data frame
data(dataCar)

# Load the tabplot package
suppressPackageStartupMessages(library(tabplot))

p<-tablePrepare(dataCar) #Dont include in datacamp

# Generate the default tableplot() display
#tableplot(p)   --- this needs work
```

## A lattice graphics example
Another package built on grid graphics is lattice, but unlike tabplot, lattice is a general-purpose graphics package that provides alternative implementations of many of the plotting functions available in base graphics. Specific examples include scatterplots with the xyplot() function, bar charts with the barchart() function, and boxplots with the bwplot() function.

One important difference between lattice graphics and base graphics is that similar functions available in both graphics systems often produce very different results when applied to the same data. As a specific example, the bwplot() function creates horizontal boxplots, while the default result of the boxplot() is a vertical boxplot display.

Another more important difference between lattice and base graphics is that lattice graphics supports conditional plots that show the separate relationships between variables within different groups. This capability is illustrated in this exercise, where you are asked to construct a plot showing the relationship between the variables calories and sugars from the UScereal data frame, conditional on the value of the shelf variable.

Instructions
100 XP
Instructions
100 XP
Load the lattice package to make the function xyplot() available.
Create the formula of calories vs. sugars conditional on shelf from the UScereal data frame in the MASS package. It takes the form y ~ x | z. Assign this as calories_vs_sugars_by_shelf.
Draw a conditional scatterplot using xyplot(), passing the formula and the dataset.

#### Hint

Hint
Use the formula calories ~ sugars | shelf.
xyplot() takes two arguments: the x argument is the formula, calories_vs_sugars_by_shelf and data is UScereal.

```{r}
# Load the lattice package
library(lattice)

# Construct the formula
calories_vs_sugars_by_shelf <- calories ~ sugars | shelf

# Use xyplot() to draw the conditional scatterplot
xyplot(calories_vs_sugars_by_shelf, data = UScereal)
```

## A ggplot2 graphics example
This final exercise provides an introduction to the ggplot2 graphics package. Like the lattice package, the ggplot2 package is also based on grid graphics and it also represents a general purpose graphics package.

The unique feature of the ggplot2 package is that it is based on the grammar of graphics, a systematic approach to building and modifying graphical displays, starting with a basic graphics element and refining it through the addition of successive modifiers.

This exercise provides a simple illustration of the approach. Specifically, you are asked to use ggplot2 to, first, create a simple scatterplot of MPG.city versus Horsepower from the Cars93 data frame. Next, you are asked to add a simple modifier that adds color based on the value of the Cylinders variable. Finally, you are asked to convert this result into a colored bubble plot, with both bubble sizes and colors determined by the Cylinders variable.

The primary purpose of this exercise is to give you a flavor of the ggplot2 package. DataCamp offers several courses dedicated to using this popular graphics package to generate high-quality data visualizations.

Instructions
100 XP
Instructions
100 XP
The character vector IScolors from the previous exercises is still available in your workspace.

Load the ggplot2 package.
Create basePlot as the ggplot object based on the Cars93 data frame from the MASS package and the scatterplot aesthetic with x as the Horsepower variable and y as the MPG.city variable. When passing the variable names to x and y, they should be unquoted (e.g. x = var_name). (Note that this definition does not render the scatterplot.)
Make a simple rendering of basePlot with the geom_point() function.
Make a second rendering using the color parameter of geom_point() to specify the first six Iliinsky-Steele colors for the six Cylinders levels.
Make a third rendering with the same coloring as the second one, but with the point sizes corresponding to the levels of the Cylinders factor variable.

#### Hint

Hint
Note the general sequence is: (1) define basePlot; (2) render the basic plot as basePlot + geom_point(); (3) Enhance the plot by using optional arguments to the geom_point() function.
To define basePlot, the aes() layer should be called inside the ggplot() function like this:
ggplot(Cars93, aes(x = Horsepower, y = MPG.city))
Point sizes are specified in geom_point() with the size argument

```{r}
# Load the ggplot2 package
library(ggplot2)

# Create the basic plot (not displayed): basePlot
basePlot <- ggplot(Cars93, aes(x = Horsepower, y = MPG.city))

# Display the basic scatterplot
basePlot + 
  geom_point()

# Color the points by Cylinders value
basePlot + 
  geom_point(color = IScolors[Cars93$Cylinders])

# Make the point sizes also vary with Cylinders value
basePlot + 
  geom_point(color = IScolors[Cars93$Cylinders], 
             size = as.numeric(Cars93$Cylinders))
```

